---
title: "AAMC_Data_Plotter"
author: "Daniel Brock"
date: "2024-03-03"
output: html_document
---

# Importing required packages

```{r setup, include=FALSE}
# Packages
library(tidyverse)
library(readxl)
library(writexl)
library(stringr)

# Working directory
cwd = getwd()
```



# Importing the required files for number of applicants over time

```{r}
# Importing and cleaning the MD data 
mdtotals <- read_xlsx(path = paste0(cwd, "/merged files/AAMC A-1 Totals Merged.xlsx"))
mdtotals$degree <- "MD"
md_app_baseline <- mdtotals[mdtotals$cycle_year == 2012, ]$applicants
mdtotals$percent_applicants_change <- ((mdtotals$applicants - md_app_baseline) / md_app_baseline) * 100
md_mat_baseline <- mdtotals[mdtotals$cycle_year == 2012, ]$matriculants
mdtotals$percent_matriculants_change <- ((mdtotals$matriculants - md_mat_baseline) / md_mat_baseline) * 100

# Importing and cleaning the MD/PhD data 
mdphdtotals <- read_xlsx(path = paste0(cwd, "/merged files/AAMC B-8 Totals Merged.xlsx"))
mdphdtotals$degree <- "MD/PhD"
mdphd_app_baseline <- mdphdtotals[mdphdtotals$cycle_year == 2012, ]$applicants
mdphdtotals$percent_applicants_change <- ((mdphdtotals$applicants - mdphd_app_baseline) / mdphd_app_baseline) * 100
mdphd_mat_baseline <- mdphdtotals[mdphdtotals$cycle_year == 2012, ]$matriculants
mdphdtotals$percent_matriculants_change <- ((mdphdtotals$matriculants - mdphd_mat_baseline) / mdphd_mat_baseline) * 100

# Merging and tidying
tots <- dplyr::bind_rows(mdtotals, mdphdtotals)
tots <- tots %>% pivot_longer(cols = c(percent_applicants_change, percent_matriculants_change), names_to = "percent_columns", values_to = "percent_change")
tots[c("per", "app_status", "chan")] = str_split_fixed(string = tots$percent_columns, pattern = "_", 3)
```

```{r}
# Graphing percent change in applicants/matriculants over time for MD and MD/PhD schools
tots_plot <- tots %>% ggplot(aes(x = cycle_year, y = percent_change, color = degree)) + 
  geom_line() + 
  geom_point(size = 2) + 
  facet_wrap(~app_status, ncol = 2) + 
  scale_color_manual(values = c("#01016f", "#d8031c")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  labs(x = "Year", y = "Percent Change Relative to 2012", color = "Degree", title = "Trends in MD & MD/PhD Applicants and Matriculants") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
tots_plot
 
# Exporting the graph
#ggsave(plot = tots_plot, filename = paste0(cwd, "/plots/totals_AAMC.pdf"), width = 2300, height = 1000, dpi = 300, units = "px")
```

```{r}
tots_plot2 <- tots %>% 
  ggplot(aes(x = cycle_year, y = percent_change, color = degree, shape = app_status, linetype = factor(app_status))) + 
  geom_line() + 
  geom_point(size = 2) + 
  scale_linetype_manual(values = c("dashed", "solid")) + 
  scale_shape_manual(values = c(1, 19)) +
  scale_color_manual(values = c("#004fcf", "#bd0202")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent Change Relative to 2012", color = "Degree Option", shape = "Application Status", title = "AAMC Admissions Data") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
tots_plot2

# Exporting the graph
#ggsave(plot = tots_plot2, filename = paste0(cwd, "/plots/totals2_AAMC.pdf"), width = 1500, height = 1000, dpi = 300, units = "px")
```

```{r}
tots_plot3 <- tots %>% 
  dplyr::filter(app_status == "applicants") %>% 
  ggplot(aes(x = cycle_year, y = percent_change, color = degree)) + 
  geom_line() + 
  geom_point(size = 2) +
  scale_color_manual(values = c("#004fcf", "#bd0202")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent Change Relative to 2012", color = "Degree", title = "AAMC Applicants Data") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
tots_plot3

# Exporting the graph
ggsave(plot = tots_plot3, filename = paste0(cwd, "/plots/totals3_AAMC.pdf"), width = 1350, height = 1000, dpi = 300, units = "px")
```




# Importing required files for male/female data 

```{r}
# Male Female percents
a11 <- read_xlsx(paste0(cwd, "/merged files/AAMC A-1 Merged.xlsx"))
a11$degree <- "MD"
b8 <- read_xlsx(paste0(cwd, "/merged files/AAMC B-8 Merged.xlsx"))
b8$degree <- "MD/PhD"
mf <- dplyr::bind_rows(a11, b8)  #merge MD-only and MD/PhD 

# Getting raw numbers of applicants from percents
mf$women_number_applicants <- round((mf$`women applicants` / 100) * mf$applications, 0)
mf$men_number_applicants <- round((mf$`men applicants` / 100) * mf$applications, 0)
mf$women_number_matriculants <- round((mf$`women matriculants` / 100) * mf$matriculants, 0)
mf$men_number_matriculants <- round((mf$`men matriculants` / 100) * mf$matriculants, 0)
```

```{r}
# Calculating percent male & female for each degree, year, and APSA region
mf_gb <- mf %>% dplyr::group_by(degree, cycle_year, APSA_region) %>% 
  summarize(
    women_sum_applicants = sum(women_number_applicants),
    men_sum_applicants = sum(men_number_applicants),
    women_sum_matriculants = sum(women_number_matriculants),
    men_sum_matriculants = sum(men_number_matriculants)
  )
mf_gb$male_applicant_percent <- (mf_gb$men_sum_applicants / (mf_gb$men_sum_applicants + mf_gb$women_sum_applicants)) * 100
mf_gb$female_applicant_percent <- (mf_gb$women_sum_applicants / (mf_gb$men_sum_applicants + mf_gb$women_sum_applicants)) * 100
mf_gb$male_matriculant_percent <- (mf_gb$men_sum_matriculants / (mf_gb$men_sum_matriculants + mf_gb$women_sum_matriculants)) * 100
mf_gb$female_matriculant_percent <- (mf_gb$women_sum_matriculants / (mf_gb$men_sum_matriculants + mf_gb$women_sum_matriculants)) * 100

# Pivoting to graph in ggplot
mf_gb <- mf_gb %>% pivot_longer(cols = c(male_applicant_percent, female_applicant_percent, male_matriculant_percent, female_matriculant_percent), names_to = "app_mat", values_to = "percent")
mf_gb[c("sex", "app_status", "per")] <- str_split_fixed(string = mf_gb$app_mat, pattern = "_", 3)
mf_gb$APSA_region_sex <- paste0(mf_gb$APSA_region, " ", mf_gb$sex)
mf_gb$APSA_region_sex <- factor(x = mf_gb$APSA_region_sex, levels = c("Mid-Atlantic male", "Midwest male", "Northeast male", "South male", "West male", "Mid-Atlantic female", "Midwest female", "Northeast female", "South female", "West female"))
mf_gb$APSA_region <- factor(x = mf_gb$APSA_region, levels = c("Mid-Atlantic", "Midwest", "Northeast", "West", "South"))
```

```{r}
# Graphing MD and MD/PhD
mf_plot <- mf_gb %>% ggplot(aes(x = cycle_year, y = percent, color = APSA_region_sex, shape = sex)) + 
  geom_line() + 
  geom_point(size = 2) + 
  facet_wrap(~degree+app_status, ncol = 2, nrow = 2) + 
  scale_color_manual(values = c("#7cb9e8", "#00308f", "#72a0c1", "#00ffff", "#7fffd4", 
                                "#922724", "#f88379", "#800000", "#ff00ff", "#c04000")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  labs(x = "Year", y = "Percent", color = "APSA Region (sex)", shape = "Sex", title = "Trends in Male and Female Applicants/Matriculants over Time") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
mf_plot

# Exporting the graph
#ggsave(plot = mf_plot, filename = paste0(cwd, "/plots/male_female_AAMC.png"), width = 2500, height = 1600, dpi = 300, units = "px")
```

```{r}
mf_gb2 <- mf_gb %>% 
  dplyr::filter(degree == "MD/PhD", sex == "female") %>% 
  mutate(
    southern_status = case_when(
      APSA_region == "South" ~ "South",
      APSA_region != "South" ~ "Other Regions"
    )
    )
mf_gb2$southern_status <- factor(x = mf_gb2$southern_status, levels = c("Other Regions", "South"))

standard_error <- function(x) {
  std_err = sd(x) / sqrt(length(x))
  return(std_err)
}

mf_gb2 <- mf_gb2 %>% group_by(southern_status, app_status, cycle_year) %>% 
  summarise(
    mean_percent = mean(percent),
    sem_percent = standard_error(percent)
  )
```

```{r}
# Graphing MD/PhD only 
mf_plot3 <- mf_gb2 %>% 
  ggplot(aes(x = cycle_year, y = mean_percent, ymin = mean_percent - sem_percent, ymax = mean_percent + sem_percent, color = southern_status, shape = app_status, linetype = factor(app_status))) + 
  geom_errorbar(width = 0.2, linetype = "solid") +
  geom_line() + 
  geom_point(size = 2) + 
  scale_linetype_manual(values = c("dashed", "solid")) + 
  scale_shape_manual(values = c(1, 19)) +
  scale_color_manual(values = c("#d1d1d1", "#bd0202")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent Female", color = "Region", shape = "Application Status", title = "MD/PhD Female Admissions") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
mf_plot3

# Exporting the graph
#ggsave(plot = mf_plot3, filename = paste0(cwd, "/plots/female_percent_AAMC.pdf"), width = 1500, height = 1000, dpi = 300, units = "px")
```



# Importing the data for race/ethicity per region and applicants/matriculants over time

```{r}
# MD Applicants
mdapps <- read_xlsx(path = paste0(cwd, "/merged files/AAMC A-10 Merged.xlsx"))
mdapps$app_status <- "Applicants"
mdapps$degree <- "MD"

# MD Matriculants
mdmats <- read_xlsx(path = paste0(cwd, "/merged files/AAMC A-11 Merged.xlsx"))
mdmats$app_status <- "Matriculants"
mdmats$degree <- "MD"

# MD/PhD Applicants
mdphdapps <- read_xlsx(path = paste0(cwd, "/merged files/AAMC B-7 Merged.xlsx"))
mdphdapps$app_status <- "Applicants"
mdphdapps$degree <- "MD/PhD"

# MD/PhD Matriculants
mdphdmats <- read_xlsx(path = paste0(cwd, "/merged files/AAMC B-9 Merged.xlsx"))
mdphdmats$app_status <- "Matriculants"
mdphdmats$degree <- "MD/PhD"

# Merging into a single file and tidying
race <- dplyr::bind_rows(mdapps, mdmats, mdphdapps, mdphdmats)
race$URMs <- race$`American Indian or Alaska Native` + race$`Black or African American` + race$`Hispanic, Latino, or of Spanish Origin` + race$`Native Hawaiian or Other Pacific Islander`
race <- race %>% dplyr::group_by(degree, app_status, APSA_region, year) %>% 
  summarise(
    sum_URMs = sum(URMs),
    sum_natives = sum(`American Indian or Alaska Native`),
    sum_black = sum(`Black or African American`),
    sum_hispanic = sum(`Hispanic, Latino, or of Spanish Origin`),
    sum_islander = sum(`Native Hawaiian or Other Pacific Islander`),
    sum_white = sum(White),
    sum_asian = sum(Asian),
    sum_other = sum(Other),
    sum_multi = sum(`Multiple Race/Ethnicity`),
    sum_unk = sum(`Unknown Race/Ethnicity`),
    sum_notusa = sum(`Non-U.S. Citizen and Non-Permanent Resident`),
    sum_tots = sum(Total)
  )
race$percent_urms <- (race$sum_URMs / race$sum_tots) * 100
```

```{r}
# Making a ggplot 
race_plot <- race %>% ggplot(aes(x = year, y = percent_urms, color = app_status)) + 
  geom_line() + 
  geom_point(size = 2) + 
  facet_wrap(~degree+APSA_region, ncol = 5, nrow = 2) + 
  scale_color_manual(values = c("#5a5a5a", "#d8031c")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  labs(x = "Year", y = "Percent URMs", color = "Application Status", title = "Percent Underrepresented in Medicine (URM) per Region") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
race_plot
 
# Exporting the graph
#ggsave(plot = race_plot, filename = paste0(cwd, "/plots/URMs_AAMC.png"), width = 4200, height = 1900, dpi = 300, units = "px")
```

```{r}
# Graphing
race_plot2 <- race %>% ggplot(aes(x = year, y = percent_urms, color = APSA_region)) + 
  geom_line() + 
  geom_point(size = 2) + 
  facet_wrap(~degree+app_status, ncol = 2, nrow = 2) + 
  scale_color_manual(values = c("#16537e", "#999999", "#6aa84f", "#f44336", "#f6b26b")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  labs(x = "Year", y = "Percent URMs", color = "APSA Region", title = "Percent Underrepresented in Medicine (URM) per Region") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
race_plot2

# Exporting the graph
#ggsave(plot = race_plot2, filename = paste0(cwd, "/plots/URMs2_AAMC.pdf"), width = 2500, height = 1600, dpi = 300, units = "px")
```

```{r}
# Graphing
race_plot3 <- race %>% dplyr::filter(degree == "MD/PhD") %>% 
  ggplot(aes(x = year, y = percent_urms, color = APSA_region)) + 
  geom_line() + 
  geom_point(size = 2) + 
  facet_wrap(~app_status, ncol = 2) + 
  scale_color_manual(values = c("#16537e", "#999999", "#6aa84f", "#f44336", "#f6b26b")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_bw() + 
  labs(x = "Year", y = "Percent URMs", color = "APSA Region", title = "Percent Underrepresented in Medicine (URM)") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
race_plot3

# Exporting the graph
#ggsave(plot = race_plot3, filename = paste0(cwd, "/plots/URMs3_AAMC.png"), width = 2300, height = 1000, dpi = 300, units = "px")
```

```{r}
race2 <- race %>% 
  dplyr::filter(degree == "MD/PhD") %>% 
  mutate(
    southern_status = case_when(
      APSA_region == "South" ~ "South",
      APSA_region != "South" ~ "Other Regions"
    )
    )
race2$southern_status <- factor(x = race2$southern_status, levels = c("Other Regions", "South"))

standard_error <- function(x) {
  std_err = sd(x) / sqrt(length(x))
  return(std_err)
}

race2 <- race2 %>% group_by(southern_status, app_status, year) %>% 
  summarise(
    mean_percent = mean(percent_urms),
    sem_percent = standard_error(percent_urms)
  )
```

```{r}
# Graphing MD/PhD only 
race_plot4 <- race2 %>% 
  ggplot(aes(x = year, y = mean_percent, ymin = mean_percent - sem_percent, ymax = mean_percent + sem_percent, color = southern_status, shape = app_status, linetype = factor(app_status))) + 
  geom_errorbar(width = 0.2, linetype = "solid") +
  geom_line() + 
  geom_point(size = 2) + 
  scale_linetype_manual(values = c("dashed", "solid")) + 
  scale_shape_manual(values = c(1, 19)) +
  scale_color_manual(values = c("#d1d1d1", "#bd0202")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent URM", color = "Region", shape = "Application Status", title = "MD/PhD URM Admissions") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
race_plot4

# Exporting the graph
#ggsave(plot = race_plot4, filename = paste0(cwd, "/plots/URMs4_AAMC.pdf"), width = 1500, height = 1000, dpi = 300, units = "px")
```



# Making a combined race gender plot for the south region

```{r}
# Filtering for only females in the south
mf_gb3 <- mf_gb %>% dplyr::filter(sex == "female", app_status == "applicant", degree == "MD/PhD", APSA_region == "South")
mf_gb3$demo <- "Female Applicants"
race3 <- race %>% dplyr::filter(app_status == "Applicants", degree == "MD/PhD", APSA_region == "South")
race3$demo <- "URM Applicants"
colnames(race3) <- c("degree", "app_status", "APSA_region", "cycle_year", "sum_URMs", "sum_natives", "sum_black", "sum_hispanic", "sum_islander", "sum_white", "sum_asian", "sum_other", "sum_multi", "sum_unk", "sum_notusa", "sum_tots", "percent", "demo") #need to change names before merge

# Merging
ferace <- dplyr::bind_rows(mf_gb3, race3)
ferace$demo <- factor(x = ferace$demo, levels = c("Female Applicants", "URM Applicants"))
```

```{r}
# Graphing female and URM MD/PhD applicants over time
ferace_plot <- ferace %>% 
  ggplot(aes(x = cycle_year, y = percent, color = demo)) + 
  geom_line() + 
  geom_point(size = 2) + 
  scale_color_manual(values = c("#bd0202", "#004fcf")) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent Applicants", color = "Demographics", title = "Female and URM Applicants to South MD/PhD Programs") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
ferace_plot

# Exporting the graph
#ggsave(plot = ferace_plot, filename = paste0(cwd, "/plots/Female_Race1.pdf"), width = 1500, height = 1000, dpi = 300, units = "px")
```

# Making a South region and non-south region plot of female and URM applicants

```{r}
mf_gb4 <- mf_gb %>% dplyr::filter(sex == "female", app_status == "applicant", degree == "MD/PhD") %>% 
  mutate(
    southern_status = case_when(
      APSA_region == "South" ~ "South Female",
      APSA_region != "South" ~ "Other-Regions Female"
    )
    )

standard_error <- function(x) {
  std_err = sd(x) / sqrt(length(x))
  return(std_err)
}

mf_gb4 <- mf_gb4 %>% group_by(southern_status, cycle_year) %>% 
  summarise(
    mean_percent = mean(percent),
    sem_percent = standard_error(percent)
  )

race4 <- race %>% dplyr::filter(app_status == "Applicants", degree == "MD/PhD") %>% 
  mutate(
    southern_status = case_when(
      APSA_region == "South" ~ "South URM",
      APSA_region != "South" ~ "Other-Regions URM"
    )
  )

race4 <- race4 %>% group_by(southern_status, year) %>% 
  summarise(
    mean_percent = mean(percent_urms),
    sem_percent = standard_error(percent_urms)
  )

colnames(race4) <- c("southern_status", "cycle_year", "mean_percent", "sem_percent")

# Merging
ferace2 <- dplyr::bind_rows(mf_gb4, race4)
ferace2[c("Region", "Demographics")] = str_split_fixed(string = ferace2$southern_status, pattern = " ", 2)
ferace2$southern_status <- factor(x = ferace2$southern_status, levels = c("South Female", "Other-Regions Female", "South URM", "Other-Regions URM"))
ferace2$Region <- factor(x = ferace2$Region, levels = c("South", "Other-Regions"))
ferace2$Demographics <- factor(x = ferace2$Demographics, levels = c("Female", "URM"))
```

```{r}
# Graphing Female and URM Applicants to MD/PhD programs and Highlighting the SOUTH
ferace_plot2 <- ferace2 %>% 
  ggplot(aes(x = cycle_year, y = mean_percent, ymin = mean_percent - sem_percent, ymax = mean_percent + sem_percent, color = southern_status, shape = Demographics)) + 
  geom_errorbar(width = 0.2, linetype = "solid") +
  geom_line() + 
  geom_point(size = 2) + 
  #ylim(0, 55) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) + 
  scale_y_continuous(limits = c(0, 52), breaks = seq(0, 52, by = 10)) +
  scale_color_manual(values = c("#bd0202", "#feb3b3", "#01016f", "#b3b3fe")) + 
  theme_classic() + 
  labs(x = "Year", y = "Percent Applicants", color = "Regions", title = "Female and URM Applicants to MD/PhD Programs") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))
ferace_plot2

# Exporting the graph
ggsave(plot = ferace_plot2, filename = paste0(cwd, "/plots/Female_Race2.pdf"), width = 1600, height = 1000, dpi = 300, units = "px")
```
























